---
title: "Midterm - Data Modeling"
author: "Aaron Rockwell"
date: "10/19/2019"
output: pdf_document
---

```{r packages backup}
#install.packages("plyr")
#install.packages("ggplot2")
#install.packages("scatterplot3d")
#install.packages("MASS")
#install.packages("faraway")
#install.packages("caTools")

library("caTools")
library("faraway")
library("knitr")
library("plyr")
library("ggplot2")
library("scatterplot3d")
library("MASS")
```

## Problem 1

# come back to problem one, do not think these are as hashed out as they could be. problem one is for sure wrong
#it should be a linear PDF because it will equal the average.
#double check problem 2 b that I need to divide by 2 on the spread, I think so though because of midterm practice
#double check on the problem with variance if the -3 is significant

Let , $Y_i =  \beta_0 + \varepsilon_i$ where  $\epsilon_i$ is independently identically normally distributed with  $N\left(\lambda,\sigma^2\right)N(\lambda,\sigma^2)$ where $\lambda$ is a positive constant. (10 pts)

A) Write the likelihood function (5 pts)

$$L =  \prod_{i=1}^n p(y_i|x_i;\beta_0,\sigma^2) $$

$$L =  \prod_{i=1}^n \frac{1}{\sqrt{2\pi\sigma^2}}exp\left(-\frac{(Y_i-\beta_0)^2}{2\sigma^2}\right) $$


B)  Find the MLE estimators for $\beta_0$  and $\sigma^2$ (5 pts)


$$Log L = -\frac{n}{2}log2\pi-\frac{n}{2}log\sigma^2 -\frac{1}{2\sigma^2} \sum_{i=1}^n (Y_i-\beta_0)^2 $$
$$ \frac{\partial L}{\partial \beta_1}=\sum_{i=1}^n (Y_i-(\beta_{0})^2)=0 $$
$$ \frac{\partial L}{\partial \sigma}=-\frac{n}{\sigma}+\frac{1}{\sigma^3}\sum_{i=1}^n (Y_i-\beta_{0})^2=0 $$
$$ \sigma^2=\frac{\sum_{i=1}^n (Y_i-\beta_0)^2}{n}$$

\newpage
## Problem 2

Refer to the question 2 data, (40 pts)

A- ) Fit a regression model to predict Y. Obtain, the residuals and plot them against the fitted values. Also prepare a normal probability plot. What do your plots show? (5pts)

B- ) Calculate the simultaneous 90% confidence interval for (assumed: $b_0~and~b_1$) (5pts)
 
C- ) Calculate the simultaneous 90% confidence intervals for the predicted new X values for 85 and 90. (5 pts)

D- ) Conduct the Brown-Forsythe test to determine whether or not the error variance varies with the level of X. (10 pts)

E- ) Use the Box-Cox procedure to find an appropriate power transformation and perform the transformation. Obtain, the residuals and plot them against the fitted values. Also prepare a normal probability plot. What do your plots show? (15 pts)


```{r problem 2}

question2.df = read.csv("question2.csv")

question2.reg = lm(y~x, data = question2.df)

#plot(question2.reg)

par(mfrow=c(1,2))
plot(question2.reg$fitted.values,question2.reg$residuals,xlab="Residuals",ylab="Fitted Values")
qqnorm(question2.reg$residuals, ylab="Standardized Residuals",xlab="Normal Scores", main="QQ Plot") 
qqline(question2.reg$residuals, col = "steelblue", lwd = 2)


```
## Problem 2 a answer
The QQ plot shows that there is non-linearity, also the residuals vs fitted values show an uneven distribution, indicating unequal variance.


B- ) Calculate the simultaneous 90% confidence interval for (assumed: $b_0~and~b_1$) (5pts)

```{r problem 2 part b}
confint(question2.reg,level=1-0.05/2)
question2.reg

```
## Problem 2 b answer
$b_0$ = 923.85932 to 1478.38940
$b_1$ = 37.08916 to 58.00822



C- ) Calculate the simultaneous 90% confidence intervals for the predicted new X values for 85 and 90. (5 pts)

```{r problem 2 part c}

Xh<-c(85,90)
predict.lm(question2.reg, data.frame(x  = c(Xh)),interval = "confidence", level = 1-0.1/2)


#Remember there are only TWO (2) values for the interval AND 90%, need to look this up to be sure
#confidence, not predict
```
## Problem 2 c answer
X=85, $\hat{Y}$ = 4664.317 to 5821.210
X=90, $\hat{Y}$ = 4857.267 to 6103.746


D- ) Conduct the Brown-Forsythe test to determine whether or not the error variance varies with the level of X. (10 pts)


```{r problem 2 part d}
ei = question2.reg$residuals

DM = data.frame(cbind(question2.df$y,question2.df$x,ei))

M = median(question2.df$x)

DM = data.frame(cbind(question2.df$y,question2.df$x,ei))

DM1 = DM[DM[,2]< M,]
DM2 = DM[DM[,2]>=M,]

M1 = median(DM1[,3])
M2 = median(DM2[,3])

N1 = length(DM1[,3])
N2 = length(DM2[,3])

d1 = abs(DM1[,3]-M1)
d2 = abs(DM2[,3]-M2)
s2 = sqrt((var(d1)*(N1-1)+var(d2)*(N2-1))/(N1+N2-2))
Den = s2*sqrt(1/N1+1/N2)
Num = mean(d1)-mean(d2)
T = Num/Den
T
```
## Problem 2 d answer
The t stat is greater than 1 (at 3.267609), indicates that the null should be rejected. There is unequal variance.


E- ) Use the Box-Cox procedure to find an appropriate power transformation and perform the transformation. Obtain, the residuals and plot them against the fitted values. Also prepare a normal probability plot. What do your plots show? (15 pts)

```{r problem 3 part e}

#par(mfrow=c(1,3))

boxcox(question2.reg,seq(0.2,.5,by=0.05))

question2.transf.df = question2.df

question2.transf.df$y = (question2.transf.df$y)^.32

question2.transf.reg = lm(question2.transf.df$y~question2.transf.df$x)

#plot(question2.transf.reg)

plot(question2.transf.reg$fitted.values,question2.transf.reg$residuals,xlab="Residuals",ylab="Fitted Values")
qqnorm(question2.transf.reg$residuals, ylab="Standardized Residuals",xlab="Normal Scores", main="QQ Plot") 
qqline(question2.transf.reg$residuals, col = "steelblue", lwd = 2)

```
## Problem 2 e answer
The Box-Cox shows a $\lambda$ peak at around .32, this would suggest a power transformation. Transforming Y to the power of .32. The transformation straightened the QQ plot from the previous and the residuals vs fitted has less variance.



\newpage
## Problem 3

Refer to the question 2 data. (25 pts)

A-) Create development sample and hold out sample. Development sample is a random sample of 70% of the data and hold out sample is the remainder 30% of the data. Use set.seed(1023) to select the samples. (10 pts)

B-)Build the model on the development sample. Obtain, the residuals and plot them against the fitted values. Also prepare a normal probability plot. What do your plots show? (5 pts)

C-)Calculate R Square on the hold out sample (hint: calculate SSE, SSR and SST on the hold out sample). (10 pts)


```{r problem 3 a}
set.seed(1023)

ind = sample(1:nrow(question2.df), size =nrow(question2.df)*0.70)
dev = question2.df[ind,]
holdout = question2.df[-ind,]

par(mfrow=c(1,2))
plot(dev)
plot(holdout)

```
## Problem 3 a answer
Created that samples as shown in the code above as dev and holdout. Also, plotted to show selection.




B-)Build the model on the development sample. Obtain, the residuals and plot them against the fitted values. Also prepare a normal probability plot. What do your plots show? (5 pts)

```{r problem 3 b}

dev.reg = lm(y~x,data=dev)
plot(dev.reg$fitted.values,dev.reg$residuals,xlab="Residuals",ylab="Fitted Values")
qqnorm(dev.reg$residuals, ylab="Standardized Residuals",xlab="Normal Scores", main="QQ Plot") 
qqline(dev.reg$residuals, col = "steelblue", lwd = 2)

```
## Problem 3 b answer
The plots show the same issues that effected the complete dataset, with the QQ not showing a good fit and the probability that the function would need to be transformed.



C-)Calculate R Square on the hold out sample (hint: calculate SSE, SSR and SST on the hold out sample). (10 pts)

```{r problem 3 b}
holdout.df = data.frame(cbind(holdout$y,holdout$x,predict(dev.reg, holdout)))
#holdout.df

#SSE = sum((Y - Yhat)^2)
SSE = sum((holdout.df$X1-holdout.df$X3)^2)
SSE

#SSR sum(Yi-Yaverage)
Yhat.holdout = mean(holdout.df$X1)
SSR = sum((holdout.df$X3-Yhat.holdout)^2)
SSR

#2.9

#SSTO = SSR + SSE
#SSR = 3.588
#SSE = 45.818
#SSTO = 3.588 + 45.818
#R.Squared = 1 - SSE / (SSR + SSE)
R.Squared.holdout = 1 - SSE / (SSR + SSE)
R.Squared.holdout

#summary(question2.reg)

```
## Problem 3 c answer

R-Squared for the sample set is .1654616


\newpage
## Problem 4

A-) Fit a linear regression function. Obtain, the residuals and plot them against the fitted values. Also prepare a normal probability plot. What do your plots show? (5 pts)

B-) Conduct the Breusch-Pagan test to determine whether or not the error variance varies with the level of X. Use α= .05. State the alternatives. decision rule, and conclusion. Is your conclusion consistent with your preliminary findings in part (a)? (10 pts)


```{r problem 4 part a}
question2.df = read.csv("question4.csv")

question2.reg = lm(y~x, data = question2.df)

```


\newpage
## Problem 5

```{r problem 5}


```
